\subsection{Analysis of \model{}}
% \section{Ablations \YA{Rather Analysis of SteerViT (since replacing with different backbone is not ablation, neither is scaling} }
\label{sec:ablations}

We ablate the key architectural and training decisions in \model{} to isolate the contribution of each component.
Unless stated otherwise, all experiments use DINOv2 ViT-B/14 as the frozen visual backbone and RoBERTa-Large as the text encoder, trained at $336{\times}336$ resolution with batch size 12 for 500k iterations (${\sim}$96 H100 GPU-hours) on the full data mixture (\cref{sec:data}), using AdamW with a cosine schedule that warms up to $3{\times}10^{-4}$ over 5k steps, decays to $3{\times}10^{-5}$ by 40k steps, and remains constant thereafter.
The default gate $\alpha_l = 1$\JR{shouldn't the initial gate value be 0?}.
% \MT{what is the default CA gate value during inference. also, indicate if this is pre/post tanh}\JR{it is the learned one. If we do, then we just vary the learned weights by a certain factor} \MT{not true? then why does fig 1 star appear at $\alpha_l = 1$?}\JR{since a factor of 1.0 is essentially not doing any changes -> default setup. Only if for != 1.0 we change model behavior. }
We evaluate on four metrics: fine-grained classification probe accuracy (FG-CLS; \cref{subsec:pareto}), text steerability via CORE retrieval (\cref{subsec:cond_ret_exp}), and personalized object discrimination PR-AUC on PODS with descriptive text prompts (\cref{subsec:text_density_exp}).
% \MG{should we keep class, supercat?} 
% \MT{no, at least obj name / description}


% ------------------------------------------------------------------
\paragraph{Architecture Choices.}\input{tables/arch_ablation_table}
\label{subsec:arch_ablation}


% \cref{tab:arch_ablation} isolates the contribution of each architectural component by modifying one design choice at a time from the full \model{} configuration.
\model{} has three design choices \JR{there are many more design choices -> maybe ``principal design choices''; perhaps ``dimensions/components'' is better than choices}:
early fusion of text within the ViT,
tanh-gated CA layers, and
MLP text projector.
\cref{tab:arch_ablation} ablates each component.


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\textit{Early vs.\ late fusion.}
We interleave gated cross-attention within ViT layers (early fusion).  
Late fusion (\cref{tab:arch_ablation}, row 3) adds text only after the final ViT layer.  
While late fusion achieves similar steerability (93.4) and higher FG-CLS (86.5 vs.\ 77.0\JR{either this or the number in the table is incorrect}), it reduces PODS performance by 10 points\JR{potentially incorrect} (45.3 vs.\ 55.3\JR{same here}) with descriptive prompts.  
With coarse supercategory prompts, the gap disappears, indicating early fusion is important for fine-grained vision–language modeling.

\textit{Role of gating.}
Each cross-attention block uses a zero-initialized tanh gate (\cref{eq:gated_ca}), starting training from a frozen ViT and gradually incorporating text.  
Removing gating (row 4) collapses steerability and FG-CLS by 15.1 and 52.8 points below our frozen baseline (row 1), showing that ungated cross-attention disrupts frozen features.

\textit{Language projector.}
A two-layer MLP projects RoBERTa embeddings to the ViT feature space (\cref{sec:arch}).  
Replacing it with a linear projector (row 5) lowers FG-CLS by 5.5 and PODS by 3.8 points, indicating that the MLP better aligns modalities.

% \paragraph{Early vs.\ late fusion.}
% Our method interleaves gated cross-attention blocks within the ViT layers (early fusion).
% The alternative, late fusion (row 3 of \cref{tab:arch_ablation}), appends them after the final ViT layer adding text information to the ViT output features.
% Late fusion achieves comparable steerability (93.4) and better preserves FG-CLS (86.5 vs.\ 77.0),
% % \MT{what do we train here? only adapter? then this looks pretty good to me! perhaps focus on differentiator that cannot do dense pred}
% but degrades performance on PODS by 10 points (45.3 vs.\ 55.3) when conditioning on descriptive text prompts.
% On PODS with coarse supercategory prompts, however, we find the gap between early and late fusion vanishes, indicating that early fusion is necessary for fine-grained vision-language modeling.


% \paragraph{Role of gating.}
% Each cross-attention block uses a zero-initialized tanh gate (\cref{eq:gated_ca}).
% Training starts as a frozen ViT at initialization and gradually incorporates text.
% Removing gating (row 4) collapses both steerability and FG-CLS by 28.4 and 52.8 points below the frozen baseline respectively. This confirms that unconstrained cross-attention catastrophically interferes with frozen features.

% \paragraph{Language projector.}
% A two-layer MLP projects RoBERTa embeddings into the ViT's feature space prior to cross-attention (\cref{sec:arch}).
% Replacing it with a linear projector (row 5) degrades FG-CLS by 5.5 points and PODS by 3.8 points.
% This illustrates that MLP's additional expressivity aids representation alignment, allowing the cross-attention layers to focus on contextualizing visual features with text rather than bridging the modality gap. \textbf{}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% while steerability remains comparable (94.2 vs.\ 93.4).


% \MT{this can probably be just moved to supplement}

% TODO: insert results and interpretation once numbers are ready, e.g.:
% Our default spread-out placement achieves the best PODS (XX.X),
% while restricting to late layers preserves FG-CLS but reduces steerability,
% suggesting that text conditioning in early layers is necessary for reshaping
% intermediate representations rather than merely modulating the final output.




% ------------------------------------------------------------------
\paragraph{Generalization Across Pretrained Backbones.}
\label{subsec:cross_backbone}

\input{tables/cross_backbone_table}

To validate that our approach generalizes beyond DINOv2, we apply \model{} to two additional ViTs: SigLIP~\cite{zhai2023_siglip}
% (contrastive vision-language pretraining)
and MAE~\cite{he2022mae}.
% (masked autoencoding).
\cref{tab:cross_backbone} compares vanilla, late fusion, and early fusion (\model{}) variants for each backbone.


Early fusion consistently outperforms late fusion across all three ViT families.
The advantage is most pronounced for MAE, where early fusion nearly doubles late-fusion steerability (62.6 vs.\ 35.5) and improves PODS by 5.3 points.
For SigLIP, early fusion boosts steerability by 7.4 points (90.9\JR{table says 90.8} vs.\ 83.5) and PODS by 8.3 points (23.0 vs.\ 14.7).
On MVTech anomaly segmentation, an out-of-distribution task absent from training, early fusion outperforms late fusion for every backbone, confirming that the richer multimodal representations from early fusion transfer better to novel domains.
DINOv2 produces the strongest steerable representations overall, consistent with its richer self-supervised features; SigLIP and MAE both benefit from text conditioning but start from weaker baselines.
In fact, we find that steering enriches \JR{vague} MAE's visual features, as illustrated in \cref{fig:gate_factor}.
\newline\newline
We include detailed results for anomaly segmentation,
results on other tasks for different backbones,
impact of model sizes (ViT-S/B/L, RoBERTa-L/B),
progression across training iterations, and
impact of proxy training task (segmentation vs.\ Gaussian heatmap)
in the supplementary material.

 

\begin{comment}
% ------------------------------------------------------------------
\subsection{Training Duration}
\label{subsec:training_duration}

Steerability emerges rapidly: within 100k iterations, CORE accuracy reaches 85.5 (vs.\ 43.5 for frozen DINOv2), while FG-CLS degrades only modestly to 82.8.
However, tasks that require deeper language understanding continue improving with longer training: between 100k and 500k iterations, PODS improves from 35.0 to 55.3
%(+58\%)
and RefCOCOg from 58.8 to 71.5,
%(+25\%)
even as steerability saturates by 300k iterations at 92.5\%.
This pattern holds for both early and late fusion (late-fusion PODS: 39.1 at 300k vs.\ 45.3 at 500k), suggesting that longer training instills richer multimodal representations rather than merely teaching the model to route attention.
We adopt 500k iterations for all final models.
\end{comment}






\begin{comment}
% ------------------------------------------------------------------
\subsection{Training Signal}
\label{subsec:training_signal}

% Patch-level localization admits two natural supervision strategies: \textbf{pointing}, via a Gaussian heatmap centered on the bounding-box midpoint ($\sigma{=}1.1$), and \textbf{segmenting}, via the tokenized object mask used in our objective (\cref{sec:objective}).
% Segmentation-mask supervision outperforms the Gaussian alternative across all evaluation metrics (full comparison in \cref{app:training_signal}).
% Beyond raw performance, the mask signal produces a notable emergent behavior: when multiple instances of the queried class appear in an image, the model activates on all matching regions simultaneously, despite training annotations labeling only one referent per sample.
% This arises because the mask target teaches the cross-attention layers to identify \textit{which tokens match the text} rather than \textit{where the most likely centre is}, thus any token whose content matches the query receives high activation regardless of instance count.
% The Gaussian-trained model, by contrast, collapses to a single peak, consistent with its unimodal supervision.
% We provide further discussion and qualitative examples of this emergence in \cref{app:training_signal}.
% \MT{numbers pending; can also go to supplement}

Patch-level localization admits two supervision strategies: \textbf{pointing} 
% (Gaussian heatmap at the bounding-box center, $\sigma{=}1.1$) 
and \textbf{segmenting}
% (tokenized object mask, \cref{sec:objective}).
We find that segmentation-mask supervision, at the ViT feature map level,  outperforms pointing across all metrics.
It also produces a striking emergent behavior: the mask-trained model localizes \emph{all} instances of a queried class simultaneously, despite training annotations labeling only a single referent per sample.
This arises because the mask signal encourages
% cross-attention layers 
identifying \textit{which tokens match the text} rather than \textit{where the most likely centre is}.  
We provide further discussion and qualitative examples in \cref{app:training_signal}.
\end{comment}






% ------------------------------------------------------------------
\begin{comment}
\subsection{Scaling}
\label{subsec:scaling}

% \input{tables/scaling_table}
\MT{TODO numbers. TODO description; do it briefly, perhaps as a single paragraph.}


\paragraph{Visual Encoder.}
We scale the frozen DINOv2 backbone from ViT-S through ViT-B to ViT-L, and the SigLIP backbone from ViT-B to ViT-L.
% TODO: fill in numbers and interpretation once results are ready. Expected framing:
% Larger backbones improve both FG-CLS and steerability, with DINOv2 ViT-L achieving
% XX.X steerability and XX.X PODS.
% For SigLIP, scaling from ViT-B to ViT-L yields similar gains.
% This confirms that \model{}'s cross-attention layers benefit from richer
% frozen features without requiring retraining of the adapter.

\paragraph{Text Encoder.}
We compare RoBERTa-Base (125M parameters) against RoBERTa-Large (355M) as the frozen text encoder. \MG{move to supplementary; don't have results for different roberta baseline}
% TODO: fill in numbers and interpretation once results are ready. Expected framing:
% RoBERTa-Large improves PODS by X points, consistent with the finding
% that downstream text understanding benefits from richer language representations.
% Steerability is comparable, suggesting that the basic ability to route attention
% does not require a large language model.
\end{comment}

