\section{Conclusion}

We introduce Steerable Visual Representations (\model{}), a new class of visual representations that equip \textit{any} pretrained visual encoder with 
the ability to steer its features with natural language. 
% text-steerable features.
\model{} inverts the 
% dominant multimodal 
MLLM
paradigm by conditioning the visual encoder on language.             
By interleaving lightweight cross-attention layers into frozen ViT blocks, our method steers both global and local features with text while preserving the base ViT's representation quality, achieving a Pareto improvement over prior approaches with only ${\sim}$21M trainable parameters.
% We illustrate that \model{} can reorganize its embedding space using text, controlling both the level of semantic abstraction and clustering criteria. 
Across personalized object discrimination and industrial anomaly segmentation, 
% text-based steering
our method
matches or surpasses 
% dedicated
dedicated methods without task-specific training, generalizing across multiple ViT backbones.
These results suggest that text can serve as a lightweight, post-hoc steering mechanism that extends the capabilities of rich 
% unimodal 
vision encoders to new domains without fine-tuning.

% We include additional ablations and scaling experiments in \cref{apdx:more_ablations}.

\section{Acknowledgements}
The authors gratefully acknowledge the HPC resources provided by the Erlangen National High Performance Computing Center (NHR@FAU) of the Friedrich-Alexander Universität Erlangen-Nürnberg (FAU) under the BayernKI project v115be. BayernKI funding is provided by Bavarian state authorities. The authors also gratefully acknowledge the Gauss Centre for Supercomputing e.V. (www.gauss-centre.eu) for funding this project by providing computing time on the Supercomputer JUPITER at Jülich Supercomputing Centre (JSC). MT thanks Adobe Research.